---
title: "Data exploration with World Bank Migration dataset"
author: "Jens Hau"
date: "7/4/2022"
output:
  html_document:
    df_print: "paged"
    number_sections: no
    theme: flatly
    highlight: kate
    toc: yes
    toc_float:
      collapsed: true
      smooth_scroll: true
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r, echo=FALSE, include=FALSE}
#Packages used

library(tidyverse)
library(readxl)
library(stringr)
library(janitor)
library(ggthemes)
library(plotly)
library(htmlwidgets)
library(hrbrthemes) 
```

# Loading the data set

The data was pulled from the World Bank data bank from the global bilateral\
migration data set which can be accessed [here](https://databank.worldbank.org/source/global-bilateral-migration).

```{r}
WB_Data_relevant_columns <- read_excel("C:/Users/PC/Desktop/R Working directory/WB Data, relevant columns.xlsx")

wb_data <- WB_Data_relevant_columns

head(wb_data)
```

The data as it is pulled from the data bank comes in a wide format. To aggregate the migration data for each country, the data is wrangled into a tidy format.

# Making data into tidy format

```{r warning=FALSE}
wb_tidy <- wb_data %>% gather(year, migrants, c('1960', '1970', '1980', '1990', '2000')) %>%
  mutate_at(c("migrants", "year"), as.integer) %>% drop_na()

head(wb_tidy)
```

The new data set has 255.380 rows and six columns with each row representing a bilateral emigrant stock. The drop_na() function removed 12.580 NA entries. These mostly comprised small countries and territories like Isle of Man and Sao Tome or countries that didn't exist in the 1960s like Serbia. 

The next step is to sum all the bilateral emigrant stock values for each country in order to get the total emigrant population for each country. 

To do this, a for-loop is used to capture each country name in country of origin for each decade and the sum for each conutry. The capture.output() function is used to capture the output of the for-loop but turns the calculated sums into a character vector with a space attached. The cat() function is used to remove this and belowe the summed values can be turned into an interger. 

```{r warning=FALSE}
total1960 <- capture.output(for (x in unique(wb_tidy$CountryOrigin)) {
  wb_tidy %>% filter(year == 1960 & CountryOrigin %in% x) %>% 
    pull(migrants) %>% sum(na.rm = TRUE) %>% cat("\n")
})

#Turning the output into a list of numbers: 

total21960 <- as.integer(total1960)

#repeating for each decade

total1970 <- capture.output(for (x in unique(wb_tidy$CountryOrigin)) {
  wb_tidy %>% filter(year == 1970 & CountryOrigin %in% x) %>% 
    pull(migrants) %>% sum(na.rm = TRUE) %>% cat("\n")
})

total21970 <- as.integer(total1970) 


total1980 <- capture.output(for (x in unique(wb_tidy$CountryOrigin)) {
  wb_tidy %>% filter(year == 1980 & CountryOrigin %in% x) %>% 
    pull(migrants) %>% sum(na.rm = TRUE) %>% cat("\n")
})

total21980 <- as.integer(total1980)


total1990 <- capture.output(for (x in unique(wb_tidy$CountryOrigin)) {
  wb_tidy %>% filter(year == 1990 & CountryOrigin %in% x) %>% 
    pull(migrants) %>% sum(na.rm = TRUE) %>% cat("\n")
})

total21990 <- as.integer(total1990)


total2000 <- capture.output(for (x in unique(wb_tidy$CountryOrigin)) {
  wb_tidy %>% filter(year == 2000 & CountryOrigin %in% x) %>% 
    pull(migrants) %>% sum(na.rm = TRUE) %>% cat("\n")
})

total22000 <- as.integer(total2000)

#Creating a new tibble with total migration numbers

Countryname <- unique(wb_tidy$CountryOrigin)
Totalmigration <- as_tibble(Countryname)

#adding total emigrants for each decade

Totalmigration$'1960' <- total21960
Totalmigration$'1970' <- total21970
Totalmigration$'1980' <- total21980
Totalmigration$'1990' <- total21990
Totalmigration$'2000' <- total22000


#NB: this is a wide data format. use gather to create a tidy format which then can be added
# GPP/Cap etc
head(Totalmigration)
summary(Totalmigration)
```

The newly summed data with total emigrant stocks is turned into a tidy format again so the data on GDP and population can be added to the data. The new data set has 1.130 rows and three columns. ½

```{r warning=FALSE}
Totalmigration_tidy <- Totalmigration %>% gather(year, migrants, c('1960', '1970', '1980', '1990', '2000')) %>%
  mutate_at(c("migrants", "year"), as.integer) %>% drop_na()

# correcting column name
Totalmigration_tidy <- rename(Totalmigration_tidy, country = value)

head(Totalmigration_tidy)
```

## Importing data from Penn World Tables

The data used for GDP and population size is Penn World Table version 10.0 and can be accessed [here](https://www.rug.nl/ggdc/productivity/pwt/?lang=en).

The data is chosen from Penn World Tables instead of the World Bank because there are fewer missing values. While more values for population and GPP are available, it comes at the cost of risking issues connected to comparability between the migration data and the population data. After having completed the graphs for the data, it is similar to the graphs made by Clemens and therefore looks to be justified comparison. 

the variable rgdpe is Expenditure-side real GDP at chained PPPs (in mil. 2017US$).

The data is moreover sorted to exclude countries with populations lower than 1.000.000 as these countries often exhibit very large migration to population ratios due to the small populations and they are therefore not as representative of the more general pattern. Moreover, a high migrant to population ratio from countries with small populations also present next no political dilemmas.

This results in N=91 for 1960, N=111 for 1970, N=116 for 1980, N=141 for 1990 and N=145 for 2000.

```{r warning=FALSE}
Penn_World_Tables_2022 <- read_excel("C:/Users/PC/Desktop/R Working directory/Penn World Tables 2022.xlsx")

Penn <- select(Penn_World_Tables_2022, country, year, pop, rgdpe)


#The Penn population data is in millions, so it is multiplied by 1.000.000 
#in order for to be matched with the migration data. 
Penn <- Penn %>% mutate(pop = pop*1000000)

#The data is sorted to only include countries with a population exceeding 
#one million, as countries with very small populations have an unrepresentative
#migration to population ratio.
Penn <- subset(Penn, pop > 1000000)

#adding GDP per capita
Penn <- Penn %>% mutate(gdpcap = (rgdpe*1e+06)/pop)

penn_sixty <- subset(Penn, year == 1960) %>% na.omit

penn_seventy <- subset(Penn, year == 1970) %>% na.omit

penn_eighty <- subset(Penn, year == 1980) %>% na.omit

penn_ninety <- subset(Penn, year == 1990) %>% na.omit

penn_twok <- subset(Penn, year == 2000) %>% na.omit
```


## Joining the two data sets

The data on GDP and population size from Penn World Tables is joined with the data on emigrant stocks for each decade

```{r warning=FALSE}

# Data set for 1960
Totalmigration_tidy_sixty <- subset(Totalmigration_tidy, year==1960)

Tomig_sixty <- left_join(Totalmigration_tidy_sixty, penn_sixty, by = "country") %>% na.omit

Tomig_sixty$year.y <- NULL

Tomig_sixty <- Tomig_sixty %>% rename("year" = "year.x")

Tomig_sixty <- Tomig_sixty %>% mutate(migpop=migrants/pop)

# Data set for 1970
Totalmigration_tidy_seventy <- subset(Totalmigration_tidy, year==1970)

#penn_seventy <- penn_seventy %>% rename("country"="Country")

Tomig_seventy <- left_join(Totalmigration_tidy_seventy, penn_seventy, by = "country") %>% na.omit

Tomig_seventy$year.y <- NULL

Tomig_seventy <- Tomig_seventy %>% rename("year" = "year.x")

Tomig_seventy <- Tomig_seventy %>% mutate(migpop=migrants/pop)

# Data set for 1980
Totalmigration_tidy_eighty <- subset(Totalmigration_tidy, year==1980)

#penn_eighty <- penn_eighty %>% rename("Country"="country")

Tomig_eighty <- left_join(Totalmigration_tidy_eighty, penn_eighty, by = "country") %>% na.omit

Tomig_eighty$year.y <- NULL

Tomig_eighty <- Tomig_eighty %>% rename("year" = "year.x")

Tomig_eighty <- Tomig_eighty %>% mutate(migpop=migrants/pop)


## data for 1990

Totalmigration_tidy_ninety <- subset(Totalmigration_tidy, year==1990)

Tomig_ninety <- left_join(Totalmigration_tidy_ninety, penn_ninety, by = "country") %>% na.omit

Tomig_ninety$year.y <- NULL

Tomig_ninety <- Tomig_ninety %>% rename("year" = "year.x")

Tomig_ninety <- Tomig_ninety %>% mutate(migpop=migrants/pop)


## data for 2000

Totalmigration_tidy_twok <- subset(Totalmigration_tidy, year==2000)

Tomig_twok <- left_join(Totalmigration_tidy_twok, penn_twok, by = "country") %>% na.omit

Tomig_twok$year.y <- NULL

Tomig_twok <- Tomig_twok %>% rename("year" = "year.x")

Tomig_twok <- Tomig_twok %>% mutate(migpop=migrants/pop)
```


### Loading continents

To examine regional differences in emigration rates, the regional definitions for each country as defined by the UN Statistical Division is added to the data set. The country and region list can be accessed [here](https://unstats.un.org/unsd/methodology/m49/).

```{r, echo = FALSE, warning=FALSE}

library(stringr)

region <- read_xlsx("C:/Users/PC/Desktop/R Working directory/Continent_list2.xlsx")

subregion <- region %>% select(country, subregion, region)

subregion <- subregion %>% mutate(country = str_trim(country))

subregion$country[80] <- "France"

subregion$country[234] <- "United Kingdom"
 
subregion$country[237] <- "United States"

subregion$country[151] <- "Myanmar"

subregion$country[61] <- "Czech Republic"

Tomig_sixty_reg <- Tomig_sixty %>% left_join(subregion, by = "country") 

#omig_sixty_reg <- Tomig_sixty_reg$rgdpe %>% round_half_up()

#view(Tomig_sixty_reg)

# 1) Think about maybe using round_half_up() with the janitor package for GDP/cap
# 2) it would be nice to have a set of code that sought out the problem rows for 
#    for countries - something like giving a list of rows with NA

### For 1970

Tomig_seventy_reg <- Tomig_seventy %>% left_join(subregion, by = "country")

#view(Tomig_seventy_reg)

### For 1980

Tomig_eighty_reg <- Tomig_eighty %>% left_join(subregion, by = "country")

#view(Tomig_eighty_reg)

### For 1990

Tomig_ninety_reg <- Tomig_ninety %>% left_join(subregion, by = "country")

#view(Tomig_ninety_reg)

### For 2000

Tomig_twok_reg <- Tomig_twok %>% left_join(subregion, by = "country")

#view(Tomig_twok_reg)

#view(subregion)

```

### Completing the data set
After having created the data for each decade with the correct calculations and labels for regions and subregions, the data is then re-combined to a final single data set

```{r warning=FALSE}
Totalemigrantstock <- bind_rows(Tomig_sixty_reg, Tomig_seventy_reg, Tomig_seventy_reg, Tomig_eighty_reg, Tomig_ninety_reg, Tomig_twok_reg)

Totalemigrantstock$region <- Totalemigrantstock$region %>% as.factor()


ggplot(Totalemigrantstock, aes(x = gdpcap, y = migpop, color = as.factor(year))) +
  geom_smooth(method = "loess", se = FALSE, span = 0.7, level = 0.5) +
  scale_x_continuous(trans = "log", breaks = c(0, 500, 5000, 50000)) +
  ylim(0,0.1) +
  #geom_point(alpha = 0.2) +
  labs(
    x = "\nGDP/capita (2017 PPP US$), log-scale",
    y = "Migrant/population\n",
    title = "Emigrant stocks in cross-section",
    color = "Year") +
  theme_minimal()
```

The final result yields a similar graphs as made by Clemens (2014) in the article "Does Development Reduce Migration?". 


